---
title: "**STA2201 Project**"
author: "Quynh Vu"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
output: pdf_document
abstract: "The unprecedented emergence of COVID-19 upended our lives to a great extent, resulting in pressing health crises and economic fallouts on a global scale. Even after a considerable effort had gone into vaccination to lift COVID-19 restrictions on our regular physical activities and stabilize outbreak incidence, there continues to be variation in the course of the COVID-19 outbreak across provinces and territories. We use countrywide and province-level data on cases and deaths during 164 consecutive weeks sourced from the [Government of Canada](https://health-infobase.canada.ca/covid-19/)  to quantify COVID-19 weekly reported new case differential among Canadian provinces and potentially forecast the risk of future outbreaks using the Hierarchical Gompertz Time Series models."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  out.width = "100%",
  fig.width = 10,
  fig.height = 6.7, 
  fig.retina = 3,
  cache = TRUE)
```

```{r echo=FALSE}
library(tinytex)
library(tidyverse)
library(here)
library(rstan)
library(bayesplot) 
library(loo) 
library(tidybayes) 
library(reshape2)
library(lubridate)
library(gridExtra)
library(rvest)
library(janitor)
library(anytime) 
library(zoo)
library(Hmisc)
library(forecast)
library(ggpubr)
library(viridis) 
library(stringr)  
library(ggmap)
#install.packages("osmdata")
library(osmdata)
library('rvest')
library(skimr)
library(visdat)
```

# Introduction

## Data

## EDA

```{r echo=FALSE}
covid_df <- read.csv("https://raw.githubusercontent.com/pquynhvu/STA2201-W2023/main/Project/data/COVID19%20cases.csv") |>           
            select(- c(X_id, Assigned_ID, Outbreak.Associated)) |>
            rename(source = Source.of.Infection,
                   age_gp = Age.Group, 
                   NBHD = Neighbourhood.Name,
                   type = Classification,
                   est_acquired_date = Episode.Date,
                   reported_date = Reported.Date,
                   gender = Client.Gender,
                   hospitalized = Ever.Hospitalized,
                   ICU = Ever.in.ICU,
                   intubated = Ever.Intubated,
                   outcome = Outcome) |>
           filter(outcome != "ACTIVE") |>
          mutate_if(is.character, ~na_if(., ''))
```

```{r echo=FALSE}
skim(covid_df)
vis_miss(covid_df, warn_large_data = FALSE)
```
```{r echo=FALSE}
covid_df <- covid_df |> mutate(est_acquired_date = as.Date(est_acquired_date, "%m/%d/%Y"),
                               reported_date = as.Date(reported_date, "%m/%d/%Y"),
                               age_gp = case_when(age_gp == "50 to 59 Years" ~ '50-59',
                                                  age_gp == "20 to 29 Years" ~ '20-29',
                                                  age_gp == "60 to 69 Years" ~ '60-69',
                                                  age_gp == "80 to 89 Years" ~ '80-89',
                                                  age_gp == "70 to 79 Years" ~ '70-79',
                                                  age_gp == "30 to 39 Years" ~ '30-39',
                                                  age_gp == "40 to 49 Years" ~ '40-49',
                                                  age_gp == "19 and younger" ~ '19-',
                                                  age_gp == "90 and older" ~ '90+'),
                               source = case_when(source == "Travel" ~ "travel",
                                                  source == "Community" ~ "community",
                                                  source == "Outbreaks, Congregate Settings" ~ "congregate settings",
                                                  source == "Close Contact" ~ "close contact",
                                                  source == "Household Contact" ~ "household",
                                                  source == "No Information" ~ "no information",
                                                  source == "Outbreaks, Healthcare Institutions" ~ "healthcare institutions",
                                                  source == "Outbreaks, Other Settings" ~ "others"),
                               type = case_when(type == "CONFIRMED" ~ 'confirmed', type == "PROBABLE" ~ 'probable'),
                               gender = case_when(gender == "MALE" ~ 'male', gender == "FEMALE" ~ 'female'),
                               outcome = case_when(outcome == "RESOLVED" ~ 'resolved', outcome == "FATAL" ~ 'fatal'),
                               age_fc = case_when(age_gp == "19-" ~ 1, age_gp == "20-29" ~ 2, age_gp == "30-39" ~ 3, age_gp == "40-49" ~ 4,
                                                  age_gp == "50-59" ~ 5, age_gp == "60-69" ~ 6, age_gp == "70-79" ~ 7, age_gp == "80-89" ~ 8,
                                                  age_gp == "90+" ~ 9),
                               gender_fc = case_when(gender == "female" ~ 1, gender == "male" ~ 0),
                               outcome = as.factor(recode(outcome, `1` = "fatal", `0` = "resolved")),
                               deaths = case_when(outcome == "fatal" ~ 1, outcome == "resolved" ~ 0),
                               source_fc = case_when(source == "travel" ~ 1, source == "community" ~ 2, source == "congregate settings" ~ 3, 
                                                     source == "close contact" ~ 4, source == "household" ~ 5, source == "no information" ~ 6,
                                                     source == "healthcare institutions" ~ 7, source == "others" ~ 8)) |>
                        drop_na()
```

```{r echo=FALSE}
district_page <- read_html('https://www.zipcode.com.ng/2023/01/city-of-toronto-on.html')
district <- district_page |> html_nodes("td:nth-child(1)") |> html_text()
FSA2d <- district_page |> html_nodes("tr:nth-child(7) td:nth-child(2) , .table-responsive tr:nth-child(2) td:nth-child(2)") |> html_text()
district2d <- district_page |> html_nodes("tr:nth-child(7) td:nth-child(1) , tr:nth-child(2) td:nth-child(1)") |> html_text()
postalcode2 <- data.frame(district = district2d, FSA = FSA2d) |>
               mutate(FSA = strsplit(as.character(FSA), ",")) |> 
               unnest(FSA) |>
               mutate(district = case_when(district == "East York" ~ "Toronto and East York",
                                           district == "York" ~ "Toronto and East York"))
postalcode2$FSA <- gsub(" ","",as.character(postalcode2$FSA))
                         
FSA_page <- read_html('https://www.zipdatamaps.com/en/postal-code-lists/canada/list-of-all-forward-sortation-areas-in-ontario')
FSA <- FSA_page |> html_nodes(".table-condensed a") |> html_text()
district_of_FSA <- FSA_page |> html_nodes("td+ td") |> html_text()
postalcode1 <- data.frame(district_of_FSA, FSA) 
postalcode1$district_of_FSA <- gsub("Old ","",as.character(postalcode1$district_of_FSA))
postalcode1 <- postalcode1 |> filter(district_of_FSA %in% district) |>
               rename(district = "district_of_FSA") |>
               mutate(district = case_when(district == "Scarborough" ~ "Scarborough",
                                           district == "North York" ~ "North York",
                                           district == "Toronto" ~ "Toronto and East York",
                                           district == "York" ~ "Toronto and East York",
                                           district == "Etobicoke" ~ "Etobicoke York"))

postalcode <- rbind(postalcode1, postalcode2) |> as.data.frame() |> distinct() 
ix <- which(postalcode$district == "North York" & postalcode$FSA %in% c("M3C", "M4A", "M4G", "M4P"))
postalcode <- postalcode |> slice(- ix)

hospital_page <- read_html('https://en.wikipedia.org/wiki/List_of_hospitals_in_Toronto')
hospitals <- hospital_page |> html_nodes("td:nth-child(1)") |> html_text() |> as.data.frame() |> slice_head(n = 35)
hospital_district <- hospital_page |> html_nodes("td:nth-child(3)") |> html_text()
hospitals <- data.frame(hospitals, hospital_district) |>
             rename(hospital = "html_text.html_nodes.hospital_page...td.nth.child.1....", district = "hospital_district") |>
             mutate(district = case_when(district == "Scarborough" ~ "Scarborough",
                                         district == "North York" ~ "North York",
                                         district == "Old Toronto" ~ "Toronto and East York",
                                         district == "East York" ~ "Toronto and East York",
                                         district == "York" ~ "Toronto and East York",
                                         district == "Etobicoke" ~ "Etobicoke York")) |>
             group_by(district) |>
             mutate(num_of_hospitals = n()) |>
             ungroup()

hospitals_by_FSA <- left_join(hospitals, postalcode, by = c("district")) |> select(- hospital) |> distinct()
covid_df <- left_join(covid_df, hospitals_by_FSA, by = "FSA") |> distinct() |>
            mutate(district_fc = case_when(district == "North York" ~ 1,
                                           district == "Toronto and East York" ~ 2,
                                           district == "Scarborough" ~ 3,
                                           district == "Etobicoke York" ~ 4)) 
```

```{r echo=FALSE}
covid_df |> group_by(district, age_gp) |>
            mutate(mu = sum(deaths)/n()) |> 
            select(district, mu) |>
            distinct() |>
            ggplot(aes(y=mu, x=age_gp)) + 
                   geom_bar(position="dodge", stat="identity") +
                   facet_wrap(~district)+
                   theme_minimal()
```

```{r echo=TRUE}
ttc <- read.csv("https://raw.githubusercontent.com/pquynhvu/STA2201-W2023/main/Project/data/ttc.csv") |>
       select(Neighbourhood, TTC.Stops, Traffic.Collisions) |>
       rename(NBHD = "Neighbourhood", num_stops = "TTC.Stops", collision = "Traffic.Collisions") |>
       mutate(NBHD = case_when(NBHD == "Mimico" ~ 'Mimico (includes Humber Bay Shores)',
                               NBHD == "Dovercourt-Wallace Emerson-Juncti" ~ 'Dovercourt-Wallace Emerson-Junction',
                               NBHD == "Cabbagetown-South St.James Town" ~ 'Cabbagetown-South St. James Town',
                               NBHD == "North St.James Town" ~ 'North St. James Town',
                               NBHD == "Briar Hill-Belgravia" ~ 'Briar Hill - Belgravia',
                               TRUE ~ as.character(NBHD)))

covid_df <- left_join(covid_df, ttc, by = "NBHD") 
```

```{r echo=FALSE}
age70plus <- c('70-79', '80-89', '80-89', '90+')
pop_df <- read.csv("https://raw.githubusercontent.com/pquynhvu/STA2201-W2023/main/Project/data/wellbeing2016.csv") |>
          select(- c("X55.years.and.over" , "X65.years.and.over", "X85.years.and.over", 
                     "Total.Population...All.Age.Groups...100..data",
                     "X0.to.04.years", "X05.to.09.years", "X10.to.14.years", "NeighbourhoodID"))|>
          pivot_longer(2:19, names_to = "age", values_to = "pop_age_NBHD") |> 
          mutate(age_gp = case_when(age == "X0.to.14.years" ~ '19-',
                                    age == "X15.to.19.years" ~ '19-',
                                    age == "X20.to.24.years" ~ '20-29',
                                    age == "X25.to.29.years" ~ '20-29',
                                    age == "X30.to.34.years" ~ '30-39',
                                    age == "X35.to.39.years" ~ '30-39',
                                    age == "X40.to.44.years" ~ '40-49',
                                    age == "X50.to.54.years" ~ '50-59',
                                    age == "X55.to.59.years" ~ '50-59',
                                    age == "X60.to.64.years" ~ '60-69',
                                    age == "X65.to.69.years" ~ '60-69',
                                    age == "X70.to.74.years" ~ '70-79',
                                    age == "X75.to.79.years" ~ '70-79',
                                    age == "X80.to.84.years" ~ '80-89',
                                    age == "X85.to.89.years" ~ '80-89',
                                    age == "X90.to.94.years" ~ '90+',
                                    age == "X95.to.99.years" ~ '90+',
                                    age == "X100.years.and.over" ~ '90+'),
                 plus70_fc = case_when(age_gp %in% age70plus ~ 'over 70', TRUE ~ 'less than 70')) |>
             select(- age) |>
             drop_na()|>
             rename(NBHD = "Neighbourhood") |>
             group_by(NBHD, age_gp)|>
             mutate(pop_age_NBHD = sum(pop_age_NBHD),
                    NBHD = case_when(NBHD == "Danforth East York" ~ 'Danforth-East York',
                                     NBHD == "Briar Hill-Belgravia" ~ 'Briar Hill - Belgravia',
                                     TRUE ~ as.character(NBHD))) |>
             ungroup() |>
             distinct()
```

```{r echo=FALSE}
covid_pop_df <- left_join(covid_df, pop_df, by = c('NBHD', 'age_gp')) |>
                group_by(district) |> 
                mutate(pop_district = sum(pop_age_NBHD), 
                       prop70plus = mean(plus70_fc == "over 70"),
                       sum_ttc_stops = sum(num_stops),
                       collision_per_stop = sum(collision)/sum(num_stops)) |> 
                ungroup() |>
                arrange(district_fc)
```

```{r echo=FALSE}
tab1 <- as.data.frame(table(covid_pop_df$district)) |> rename(district = Var1)
tab = c(0, tab1$Freq)
```

# Methods

There are two levels. Let i = 1, 2, ..., N denote individuals and j = 1, 2, 3, 4 districts. Deaths in district j follow a Poisson distribution,

$$D_j \sim Poisson(\mu_j P_j)$$

where $P_j$ is an offset in form of a population in neighborhood j (in thousands), and $\mu_j$ is a death rate due to COVID-19, calculated as the average of the latent death risk variable (or vulnerability) over individuals $N_j$ in neighborhood j:

$$\mu_j = \frac{\sum_{i = 1}^{N_j} \mu_{ij}}{N_j}$$

We model the individual latent death risk with a two-level model:

$$log(\mu_{ij}) = \alpha_0 + \sum_{k = 1}^K \alpha_k X_{ijk} + \sum_{l = 1}^L \beta_l X_{jl}$$
where $X_{ij} = {`gender`, `age_gp`, `source`} are individual-level predictors, and $Z_j = {`prop80plus`, `stop_volume`, `collision_per_stop`, `num_of_hospitals`} are district-level predictors.

```{r}
dd <- covid_pop_df |> select(id, district, mu_j) |>
                      mutate(district_fc = case_when(district == "North York" ~ '1',
                                                     district == "Toronto and East York" ~ '2',
                                                     district == "Etobicoke York" ~ '3',
                                                     district == "Scarborough" ~ '4')) 

```

# Results

# Discussion

# References