---
title: "Lab 2"
author: "Quynh Vu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  out.width = "100%",
  fig.width = 10,
  fig.height = 6.7, 
  fig.retina = 3,
  cache = FALSE)
```

```{r library}
library(opendatatoronto)
library(tidyverse)
library(stringr)
library(skimr) # EDA
library(visdat) # EDA
library(janitor)
library(lubridate)
library(ggrepel)
```

```{r}
res <- list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b") 
res <- res |> mutate(year = str_extract(name, "202.?"))
delay_2022_ids <- res |> filter(year==2022)|> select(id) |> pull() 
delay_2022 <- get_resource(delay_2022_ids)
delay_2022 <- clean_names(delay_2022) 
delay_2022 <- delay_2022 |> distinct() # subset distinct/unique rows
```

```{r}
delay_2022$line <- as.factor(delay_2022$line)
delay_2022 <- delay_2022 |> filter(line %in% c("BD", "YU", "SHP", "SRT"))

delay_2022 <- delay_2022 |> mutate(station = ifelse(str_starts(station, "ST"), 
                                   word(station, 1,2), word(station, 1)))
delay_2022$station <- as.factor(delay_2022$station)
levels(delay_2022$station)[levels(delay_2022$station) == "ST ANDREW"]  <- "ST.ANDREW"
levels(delay_2022$station)[levels(delay_2022$station) == "ST PATRICK"]  <- "ST.PATRICK"
levels(delay_2022$station)[levels(delay_2022$station) == "ST CLAIR"]  <- "ST.CLAIR"
levels(delay_2022$station)[levels(delay_2022$station) == "ST. GEORGE"]  <- "ST.GEORGE"
levels(delay_2022$station)[levels(delay_2022$station) == "ST GEORGE"]  <- "ST.GEORGE"
levels(delay_2022$station)[levels(delay_2022$station) == "YONGE/SHEPPARD"]  <- "SHEPPARD-YONGE"
levels(delay_2022$station)[levels(delay_2022$station) == "YONGE-SHEPPARD"]  <- "SHEPPARD-YONGE"
levels(delay_2022$station)[levels(delay_2022$station) == "BLOOR/DANFORTH"]  <- "BLOOR-DANFORTH"
levels(delay_2022$station)[levels(delay_2022$station) == "YONGE/UNIVERSITY"]  <- "YONGE-UNIVERSITY"
levels(delay_2022$station)[levels(delay_2022$station) == "DOWNVIEW"]  <- "DOWNSVIEW"
levels(delay_2022$station)[levels(delay_2022$station) == "QUEEN'S"]  <- "QUEENS"
levels(delay_2022$station)[levels(delay_2022$station) == "GLENCARIN"]  <- "GLENCAIRN"
levels(delay_2022$station)[levels(delay_2022$station) == "LAWRECNE"]  <- "LAWRENCE"
levels(delay_2022$station)[levels(delay_2022$station) == "SUMMER"]  <- "SUMMERHILL"
```

# 1. Using the `delay_2022` data, plot the five stations with the highest mean delays. Facet the graph by `line`

```{r echo=TRUE}
top_five_delays <- delay_2022 |> 
                              group_by(station) |> 
                              mutate(avgDelay = mean(min_delay)) |>
                              arrange(desc(avgDelay))|>
                              distinct(avgDelay)
```

After removing the observations that have non-standardized lines, we recoded station names to make them consistent, e.g. both `ST. GEORGE` and `ST GEORGE` to `ST.GEORGE` or `YONGE/UNIVERSITY` to `YONGE-UNIVERSITY`. The five stations with the highest mean delays are

```{r echo=TRUE}
head(top_five_delays, 5)
```

```{r echo=TRUE}
top5 <- delay_2022 |> filter(station %in% c("GUNN", "VIADUCT", "VMC", "SRT", "BLOOR-DANFORTH"))

top5 |> mutate(week = week(date)) |> # Get/set weeks component of a date-time
        group_by(week, line) |> 
        ggplot(aes(week, min_delay, color = station)) + 
               geom_point() + 
               geom_smooth() + 
               labs(title = "Distribution of 5 stations that have the highest delay time on average over time") +
               facet_grid(~line)
top5 |> group_by(station) |> 
        ggplot(aes(station, color = station)) +
               geom_bar() +
               labs(title = "Number of delays of 5 stations that have the highest delay time on average") +
                facet_grid(~line)

top5 |> group_by(station) |> 
        mutate(totalDelay = sum(min_delay)) |>
        ggplot(aes(y = totalDelay, x = station)) +
        geom_point() +
        labs(title = "Total amount of delay time of 5 stations that have the highest delay time on average") +
        facet_grid(~line)
```

# 2. Using the `opendatatoronto` package, download the data on mayoral campaign contributions for 2014. Hints:

  * find the ID code you need for the package you need by searching for 'campaign' in the `all_data` tibble above
    
  * you will then need to `list_package_resources` to get ID for the data file
    
**Note**: the 2014 file you will get from `get_resource` has a bunch of different campaign contributions, so just keep the data that relates to the Mayor election

```{r echo=TRUE}
contribution <- list_package_resources("f6651a40-2f52-46fc-9e04-b760c16edd5c") 
campaign <- get_resource("5b230e92-0a22-4a15-9572-0b19cc222985") 
mayor2014 <- campaign[["2_Mayor_Contributions_2014_election.xls"]]
head(mayor2014)
```

# 3. Clean up the data format (fixing the parsing issue and standardizing the column names using `janitor`)

```{r echo=TRUE}
#not_all_na <- function(x) any(!is.na(x))
#select_if(not_all_na) |> # remove columns with all NAs
names(mayor2014) <- as.matrix(mayor2014[1, ])
mayor2014 <- mayor2014[-1, ]  # make the first row the header

mayor2014 <- mayor2014 |> clean_names() |>
                          rename(contributor = contributors_name, 
                                 contributor_type = contributor_type_desc,
                                 relationship = relationship_to_candidate,
                                 representative = authorized_representative,
                                 contribution_type = contribution_type_desc,
                                 manager = president_business_manager, 
                                 services = goods_or_service_desc)
names(mayor2014)[1:5][-1] = str_sub(names(mayor2014)[-1], 14)
head(mayor2014)
```

# 4. Summarize the variables in the dataset. Are there missing values, and if so, should we be worried about them? Is every variable in the format it should be? If not, create new variable(s) that are in the right format.

**Summarize the variables in the dataset:**

```{r echo=TRUE}
skim(mayor2014)
```

There are a lot of `NA`s in variables `address` (contributor's address), `services` (types of goods or services provided in place of monetary support), `relationship` (relationship between the contributor and the candidate), `manager` (name of president business manager), and `representative` (authorized representative). All of the values in the variable `ward` are missing. 

How we deal with missing values depends on our analysis goal. For instance, if we want to investigate and compare the contribution values that individuals and cooperations supported the candidates from their favourable political party, then we should not be worried about the missing information on the six variables having the highest missing value. Also, the `postal code` variable provided more comprehensive information on the contributor's residency than the variable `address` if that topic of our interest. In case the variable we are interested in has lots of NAs, then steps such as interpolation, imputation, or adding missing indicator to encode "missingness" as a feature should be taken into consideration to obtain well-fitted models.

```{r}
mayor_2014 <- data.frame(matrix(nrow = nrow(mayor2014), ncol = 1)) # create new data frame

mayor_2014$contributor <- mayor2014$contributor          
mayor_2014$address <- mayor2014$address         
mayor_2014$postal_code <- mayor2014$postal_code
mayor_2014$amount <- as.numeric(mayor2014$amount)
mayor_2014$type <- as.factor(mayor2014$type)
mayor_2014$services <- mayor2014$services
mayor_2014$contributor_type <- as.factor(mayor2014$contributor_type)
mayor_2014$relationship <- mayor2014$relationship
mayor_2014$manager <- mayor2014$manager
mayor_2014$representative  <- mayor2014$representative
mayor_2014$candidate  <- mayor2014$candidate
mayor_2014$office  <- mayor2014$office
mayor_2014$ward  <- mayor2014$ward
mayor_2014 <- mayor_2014[ -c(1)]
```

The variable `amount` should be of numeric instead of character type. We also recorded variables `type` (type of contribution) and `contributor_type` as factors with two levels as follows:

| Variable         | Levels of Factor |
|------------------|------------------|
| `type`             | Goods/Services   |
|                  | Monetary         |
| `contributor_type` | Corporation      |
|                  | Individual       |

```{r echo=TRUE}
str(mayor_2014)
```
# 5. Visually explore the distribution of values of the contributions. What contributions are notable outliers? Do they share a similar characteristic(s)? It may be useful to plot the distribution of contributions without these outliers to get a better sense of the majority of the data. 

```{r echo=TRUE}
summary(mayor_2014$amount)
```

The maximum value of the contributions is very large compared to the mean, so we suspect some influential outliers exist in the dataset. We see that most substantial contributions come from the candidate themselves or their spouse. Some notable outliers are

```{r}
# out_ind <- which(mayor_2014$amount %in% boxplot.stats(mayor_2014$amount)$out)
out <- boxplot.stats(mayor_2014$amount)$out
dd <- mayor_2014 |> filter(amount == out)
mayor_2014 |> select(contributor, amount, type, contributor_type, relationship, 
                     candidate) |>
              filter(!is.na(relationship)) |>
              filter(amount %in% dd$amount)
```

and also

```{r}
dp <- mayor_2014 |> filter(amount > mean(amount) + 2*sd(amount)) # > 2 SD from the mean
mayor_2014 |> select(contributor, amount, type, contributor_type, relationship, 
                     candidate) |>
              filter(amount %in% dp$amount)
```

```{r}
# remove potential outliers
mayor_2014_new <- mayor_2014 |> 
        filter(amount < min(boxplot.stats(mayor_2014$amount)$out))

ggp1 <- ggplot(mayor_2014, aes(x=amount)) + 
        geom_histogram(aes(y=..density..), colour="black", fill="lightblue")+
        geom_density(alpha=.2, fill="#FF6666") +
        labs(title = "Distribution of contribution values by type",
             subtitle = "(potentially) with outliers")+
        theme(legend.position="top")
ggp2 <- ggplot(mayor_2014, aes(y=amount)) +
        labs(title = "Distribution of contribution values with outliers") +
        geom_boxplot()+
        theme(legend.position = "none")

ggp3 <- ggplot(mayor_2014_new, aes(x=amount, color = type)) + 
        geom_histogram(aes(y=..density..), colour="black", fill="lightblue")+
        geom_density(alpha=.2, fill="#FF6666") +
        labs(title = "Distribution of contribution values by type",
             subtitle = "without outliers")+
        theme(legend.position="top")
ggp4 <- ggplot(mayor_2014_new, aes(y=amount)) +
        labs(title = "Distribution of contribution values without outliers") +
        geom_boxplot()+
        theme(legend.position = "none")
require(gridExtra)
grid.arrange(ggp2, ggp1, ggp4, ggp3, nrow = 2, ncol = 2)
```

# 6. List the top five candidates in each of these categories:

```{r}
mayor_2014 <- mayor_2014 |> group_by(candidate) |> mutate(sumContr = sum(amount),
                                                          avgContr = mean(amount),
                                                          count = length(amount))
```

**Total contributions**

```{r}
byTotal <- mayor_2014 |> arrange(desc(sumContr))|>
                         distinct(sumContr)
head(byTotal, 5)
```

**Mean contribution**
    
```{r}
byMean <- mayor_2014 |> arrange(desc(avgContr))|>
                        distinct(avgContr)
head(byMean, 5)
```

**Number of contributions**

```{r}
byNumber <- mayor_2014 |> arrange(desc(count))|>
                          distinct(count)
head(byNumber, 5)
```

# 7. Repeat 6 but without contributions from the candidates themselves.

```{r}
mayor_2014_new2 <- mayor_2014 |> distinct(contributor, candidate, .keep_all = TRUE)|>
                                 group_by(candidate) |> mutate(sumContr2 = sum(amount),
                                                               avgContr2 = mean(amount),
                                                               count2 = length(amount))
```

**Total contributions**

```{r}
byTotal2 <- mayor_2014_new2 |> arrange(desc(sumContr2))|>
                               distinct(sumContr2)
head(byTotal2, 5)
```

**Mean contribution**
    
```{r}
byMean2 <- mayor_2014_new2 |> arrange(desc(avgContr2))|>
                              distinct(avgContr2)
head(byMean2, 5)
```

**Number of contributions**

```{r}
byNumber2 <- mayor_2014_new2 |> arrange(desc(count2))|>
                                distinct(count2)
head(byNumber2, 5)
```

# 8. How many contributors gave money to more than one candidate? 

184 contributors gave money to more than one candidate.

```{r echo=TRUE}
mayor_2014_new3 <- aggregate(mayor_2014_new2$contributor, 
                             by=list(mayor_2014_new2$contributor, mayor_2014_new2$candidate), 
                             FUN=length) 
colnames(mayor_2014_new3) <- c("contributor", "candidate", "x") 
dm <- as.data.frame(table(mayor_2014_new3$contributor))|> filter(Freq > 1)
colnames(dm) <- c("contributor", "number of candiates supported") 
length(dm$contributor)
```

```{r}
candidate2 <- mayor_2014_new2 |> select(contributor, candidate) |>
                                 filter(contributor %in% dm$contributor)|>
                                 distinct()
candidate2
```


